\textsc{Regression And Resampling Techniques \\ FYS-STK4155 \\ Assignment One}

### 2019-09-12 assignment_one/article/main.tex
Work work.

\usepackage[backend=biber, style=authoryear]{biblatex}
\addbibresource{../../FYS-STK4155.bib}
\usepackage[noabbrev, capitalize]{cleveref}

@@

\newcommand{\x}{\bm{x}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\X}{\bm{X}}
\newcommand{\diag}[1]{\mathrm{diag}(#1)}
\newcommand{\MSE}[1]{\mathrm{MSE}(#1)}
\newcommand{\R}[1]{\mathrm{R}^2(#1)}
\newcommand{\expect}{\mathbb{E}}
\newcommand{\OLS}{\mathrm{OLS}}
\newcommand{\ridge}{\mathrm{ridge}}
\newcommand{\lasso}{\mathrm{lasso}}
\newcommand{\python}{\textsf{Python}{}}

@@

  \chapter{Regression Techniques}

@@

  One common cost function is the one which involves the sum of squared
  residuals (or squared errors):
  \begin{equation}
    \label{eq:mse}
    \cost(\data, \beta) = \frac{1}{n}\sum_{i = 1}^n \varepsilon_i^2 \coloneqq \frac{1}{n}\sum_{i=1}^n (y_i - \hat{y}_i)^2
  \end{equation}
  where \( \hat{y}_i \coloneqq f(\x_i, \beta) \). The method involving the
  minimization of this specific cost function is known as \emph{least squares}.
  With the matrix notation from above in mind, we can also write the cost
  function as
  \begin{equation}
    \cost(\data, \beta) = \frac{1}{n} (\y - \hat{\y})^T (\y - \hat{\y}).
  \end{equation}

@@

  We have that
  \begin{align}
    \dpd{\cost(\data, \beta)}{\beta} = \X^T(\y - \X\beta), 
  \end{align}
  and equating this to zero yields the familiar \emph{normal equations}
  \begin{equation}
    \X^T \y = \X^T\X\beta.
  \end{equation}
  If the matrix \( \X^T\X \) is invertible, we may obtain the solution by
  direct numerical matrix inversion. In this case, the optimal model parameters
  are found directly by
  \begin{equation}
    \label{eq:optimal}
    \beta_{\OLS} = (\X^T\X)^{-1}\X^T\y.
  \end{equation}
  However, these matrices may be ill-conditioned when the number of equations
  are very large, and it is therefore common to apply approximate solvers for
  the inverse, for instance using low-rank approximation based on
  \textsc{SVD}-decomposition, which we will briefly turn to in the following.

@@

  Since we are attempting to estimate a true distribution based on finite
  samples, there are certain statistical properties it is worthwile discussing.
  We follow the derivation from \autocite{mehtaHighbiasLowvarianceIntroduction2019}.  Recall
  \cref{as:general_linear} which states that our true data is generated from a
  noisy model
  \begin{equation}
    y = f(\x) + \varepsilon
  \end{equation}
  with \( \varepsilon \sim \N(0, \sigma^2) \).  We select a finite sample from
  the true data, which constitutes our finite data-set \( \data \).
  Furthermore, we have a model function \( f(\data, \beta) \) which depends on
  the selected data-set and the model parameters \( \beta \). Thus, our
  cost-function \( \cost \) will depend on the specific sample taken from the
  true distribution. It is of interest to examine the expected out-of-sample
  error \( \expect_{\data, \varepsilon}[\cost(\data, \beta)] \) of the model
  function over the sampled data set \( \data \) and the noise \( \varepsilon
  \). This can be decomposed into three distinct parts,
  \emph{\textcolor{Maroon}{variance}}, \emph{\textcolor{Blue}{bias}}, and
  \emph{\textcolor{Green}{noise}}:
  \begin{equation}
    \expect_{\data, \varepsilon}[\cost(\data, \beta)] = \frac{1}{n} \textcolor{Blue}{\sum_{i=1}^n \left(f(\x_i) - \expect_{\data}\left[\hat{y}_i\right]\right)^2}  + \frac{1}{n}\textcolor{Maroon}{\sum_{i=1}^n \expect_{\data}\big[\left(\hat{y}_i - \expect_{\data} [\hat{y}_i]\right)^2\big]} + \textcolor{Green} \sigma^2.
  \end{equation}
  See \cref{ap:derivation} for the full derivation. The bias-term measures the
  deviation of the expectation of the model function \(f(\data, \beta)\) from
  the true value \( f \), while the variance-term measures the fluctuations
  arising due to using finite samples.

@@

@@

  \chapter{Implementation}

  The implementations are all written in the programming language \python{} with
  extensive use of the \python-packages \textsf{numpy} and \textsf{sklearn}.

>>>>

\usepackage[backend=biber, style=alphabetic, url=false, doi=false]{biblatex}
\addbibresource{../../FYS-STK4155.bib}
\usepackage[noabbrev, capitalize]{cleveref}

@@

\newcommand{\x}{\bm{x}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\X}{\bm{X}}
\newcommand{\diag}[1]{\mathrm{diag}(#1)}
\newcommand{\MSE}[1]{\mathrm{MSE}(#1)}
\newcommand{\R}[1]{\mathrm{R}^2(#1)}
\newcommand{\expect}{\mathbb{E}}
\newcommand{\OLS}{\mathrm{OLS}}
\newcommand{\ridge}{\mathrm{ridge}}
\newcommand{\lasso}{\mathrm{lasso}}
\newcommand{\python}{\textsf{Python}{}}
\newcommand{\sgn}{\,\mathrm{sign}{}}

@@

  \chapter{Regression Techniques}
  \label{chp:regression}

@@

  One common cost function is the one which involves the sum of squared
  residuals (or squared errors):
  \begin{equation}
    \label{eq:mse}
    \cost_\OLS(\data, \beta) = \frac{1}{n}\sum_{i = 1}^n \varepsilon_i^2 \coloneqq \frac{1}{n}\sum_{i=1}^n (y_i - \hat{y}_i)^2
  \end{equation}
  where \( \hat{y}_i \coloneqq f(\x_i, \beta) \). The method involving the
  minimization of this specific cost function is known as \emph{least squares}.
  With the matrix notation from above in mind, we can also write the cost
  function as
  \begin{equation}
    \cost_\OLS(\data, \beta) = \frac{1}{n} (\y - \hat{\y})^T (\y - \hat{\y}).
  \end{equation}

@@

  We have that
  \begin{align}
    \dpd{\cost_\OLS(\data, \beta)}{\beta} = \X^T(\y - \X\beta), 
  \end{align}
  and equating this to zero yields the familiar \emph{normal equations}
  \begin{equation}
    \X^T \y = \X^T\X\beta.
  \end{equation}
  If the matrix \( \X^T\X \) is invertible, we may obtain the solution by
  direct numerical matrix inversion. In this case, the optimal model parameters
  are found directly by
  \begin{equation}
    \label{eq:optimal}
    \beta_{\OLS} = (\X^T\X)^{-1}\X^T\y.
  \end{equation}
  However, these matrices may be ill-conditioned when the number of equations
  are very large, and it is therefore common to apply approximate solvers for
  the inverse, for instance using low-rank approximation based on
  \textsc{SVD}-decomposition, which we will briefly turn to in the following.

@@

  Since we are attempting to estimate a true distribution based on finite
  samples, there are certain statistical properties it is worthwile discussing.
  We follow the derivation from \autocite{mehtaHighbiasLowvarianceIntroduction2019}.  Recall
  \cref{as:general_linear} which states that our true data is generated from a
  noisy model
  \begin{equation}
    y = f(\x) + \varepsilon
  \end{equation}
  with \( \varepsilon \sim \N(0, \sigma^2) \).  We select a finite sample from
  the true data, which constitutes our finite data-set \( \data \).
  Furthermore, we have a model function \( f(\data, \beta) \) which depends on
  the selected data-set and the model parameters \( \beta \). Thus, our
  cost-function \( \cost \) will depend on the specific sample taken from the
  true distribution. It is of interest to examine the expected out-of-sample
  error \( \expect_{\data, \varepsilon}[\cost(\data, \beta)] \) of the model
  function over the sampled data set \( \data \) and the noise \( \varepsilon
  \). This can be decomposed into three distinct parts,
  \emph{\textcolor{Maroon}{variance}}, \emph{\textcolor{Blue}{bias}}, and
  \emph{\textcolor{Green}{noise}}:
  \begin{equation}
    \expect_{\data, \varepsilon}[\cost(\data, \beta)] = \frac{1}{n} \textcolor{Blue}{\sum_{i=1}^n \left(f(\x_i) - \expect_{\data}\left[\hat{y}_i\right]\right)^2}  + \frac{1}{n}\textcolor{Maroon}{\sum_{i=1}^n \expect_{\data}\big[\left(\hat{y}_i - \expect_{\data} [\hat{y}_i]\right)^2\big]} + \textcolor{Green} \sigma^2.
  \end{equation}
  See \cref{ap:derivation} for the full derivation. The bias-term measures the
  deviation of the expectation of the model function \(f(\data, \beta)\) from
  the true value \( f \), while the variance-term measures the fluctuations
  arising due to using different samples of the true data.

  Now, note that the equation above suggests that we should minimize both the
  variance and the bias of the model, in order to obtain the smallest
  generalization error of the model function. The variance term will be larger
  for complex models, as performing a fit on one sampled data-set will not
  neccessarily translate well to another sampled data-set. The bias term
  measures errors introduced by making certain assumptions about the relations
  between the covariates \( \X \) and the response \( \y \). In our case, we
  are assuming a linear relationship (recall \emph{linear regression}), and
  thus if the true underlying relation is non-linear, our bias will be high.
  More complex models will in general have lower bias.

  This is known as the \emph{bias-variance tradeoff} as having a highly complex
  model, we reduce the bias, yet increase the variance.  Having a simple model
  (for instance a linear model), we increase the bias, and reduce the variance.

@@

  Attempting to differentiate the Lasso cost function with respect to the model parameters, we see that
  \begin{equation}
    \dpd{\cost_\lasso(\data, \beta; \lambda)}{\beta_i} = \dpd{\cost_\OLS(\data, \beta)}{\beta_i} + \lambda \sgn(\beta_i).
  \end{equation}
  The analysis of the optimal Lasso parameters is made difficult due to the
  discontinuity of the sign function, and no closed form solution as in OLS and
  Ridge-regression can be found in the general case. Thus, numerical
  optimzation techniques are often used, for instance gradient ascent/descent.

@@

  \chapter{Implementation and methodology}

  In order to verify the implementation, we test on sampled data from the
  Franke function \cite{frankeCriticalComparisonMethods1979}. This bivariate
  function \(F(x, y) \) is defined for \( x, y \in [0, 1] \) as follows:
  \begin{align}
    \begin{split}
      F(x,y) \coloneqq &\frac{3}{4} \exp\left( -\frac{(9x - 2)^2}{4} - \frac{(9y - 2)^2}{4}\right) \\
      + &\frac{3}{4}\exp\left( -\frac{(9x + 1)^2}{49} - \frac{(9 y + 1)^2}{10}\right) \\
      + &\frac{1}{2}\exp\left(-\frac{(9x-7)^2}{4} - \frac{(9y-3)^2}{4}\right) \\
      - &\frac{1}{5}\exp\left(-(9x - 4)^2 - (9y - 7)^2\right).
    \end{split} 
  \end{align}
  We are interested in fitting a bivariate polynomial surface to this data, by
  means of the regression techniques discussed in \cref{chp:regression}. That
  is, our model function \( f(\data, \beta) \) lies in the polynomial space \(
  \Pi_{2}^d \) of bi-degree \( (d, d) \) polynomials. We are mainly interested
  in bi-quintic polynomial surfaces, and will therefore restrict our attention
  to \( d = 5 \) in the numerical experiments.

  We sample the function \( F \) both with and without noise at \( N = 5000 \) points in each
  parameter direction, yielding a matrix of function values
  \begin{align}
    \begin{split}
      \mat{Z} &= \left[F(x_i, y_j) \right]_{ij}, \\
      \mat{Z}^\star &= \left[F(x_i, y_j) + \varepsilon_{ij}\right]_{ij},
    \end{split}
  \end{align}
  where each \( \varepsilon_{ij} \sim \N(0, \sigma^2) \). The design matrix \(
  \mat{X} \) is computed using the \textsc{PolynomialFeatures}-method from the
  \textsc{sklearn} package. The dimension of the bi-quintic polynomial space is
  21, yielding a design matrix of size \( N^2 \times 21 \).

  For the real data set, we use the supplied \textsc{Tiff} data of an area
  close to Stavanger in Norway, as displayed in \cref{fig:terrain}. The
  data-set consists of approximately \( 1800 \times 3600 \) data-points. 
  \begin{figure}
    \centering
    \includegraphics[width=1\linewidth]{images/terrain.pdf}
    \caption{Real terrain data from a region close to Stavanger, Norway. For
    the regression tasks, we select data from small rectangular regions to keep
  the data-sizes manageable. The data is supplied in the file
\texttt{SRTM\_data\_1\_Norway.tif}.}
  \label{fig:terrain}
  \end{figure}

  The ordinary least squares and the Ridge regression is performed by solving
  the corresponding normal equations. We use the Lasso regression available in
  \textsc{sklearn} package.
